Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-27 paper.tex
final edits

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small structure-activity
  relationship datasets). We extracted structure-activity data for several ChEMBL assays and
  developed random forest models to predict SALI values, from pairwise combinations of molecular
  descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
  significant cliffs tended to be better than on the lesser ones. Our results indicate that the
  predictive models are able to prioritize molecules in terms of their ability to activity cliffs
  and can be used as a way to extend an observed activity landscape to identify molecules that could
  lead to significant improvements in activity. Conversely, the method can be used to deprioritize
  molecules that are predicted to show an activity cliff with a potent molecule already in the
  dataset.
\end{abstract}

@@

In either case, one can take the numerical values and visualize them in a variety of ways ranging
from heatmaps of SALI matrices to network representations \cite{Wawer:2008aa,Guha:2008aa}. These
visualizations then allow the user to explore the landscape, quickly identifying a range of activity
cliffs, which an then be examined in detail. Apart from identifying individual activity cliffs, a
variety of other SAR constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and multitarget
landscapes\cite{Dimova:2011fk} can also be identified and characterized.

@@

\section{Results}
\label{sec:applications}
We first consider the application of the SALI prediction methodology to the Cavalli dataset. We
developed three random forest models, using the three aggregation descriptor aggregation functions
described in Section \ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the predicted
versus observed SALI values from the three models and Table \ref{tab:rfmodel-results} summarizes the
performance metrics for the three models. Overall the different aggregation functions do not differ
dramatically in terms of the final model performance as root mean square error (RMSE) on the test
set ranges from 1.01 to 1.07 log units. However, $f_{\textrm{mean}}$ appears to lead to the best
performance (RMSE = 1.01 and $R^2$ = 0.58). In contrast, the RMSE and $R^2$ values for
Y-scrambled\cite{Rucker:2007aa} models using the different aggregation functions ranged from 1.58 to
1.61 log units and 0.003 to 0.01, respectively.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values (for the training
set) using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{diff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics.  While the
differences in model performance do not vary significantly with the aggregation function used, we
see that for all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms
of RMSE and $R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models
exhibit more variance for small SALI values (i.e., less significant cliffs) but appear to perform
better at the higher end of the SALI values. However in all three cases, systematic prediction
errors are apparent. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$ as the
aggregation function for subsequent analyses. Table \ref{tab:rfmodel-orig} and Figure
\ref{fig:rfmodel-orig} summarizes the performance of the models built using the reported activity
data (original, single molecule activity) for the Costanzo, Kalla and Dai datasets. Each of these
models employed the entire dataset (rather than splitting into a training and testing set) and
exhibit relatively poor performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units)
and $R^2$ (0.18 to 0.34). This can be partially ascribed to the small sizes of the dataset. It
should be note that this methodology is not aimed at modeling the original, usually small, SAR
dataset. It is well known (and evident from the current analyses) that small datasets do not
necessarily lead to very robust or reliable machine learning models. Instead, the methodology
focuses on \emph{pairwise} datasets -- which are usually large enough for a statistically
significant model.

@@

\ctable[cap={}, caption={Performance metrics for the random forest
  models built for the datasets used in this study.},
label={tab:rfmodel-results}]
{lrcrrrr}
{
\tnote{Y-range is calculated using the entire dataset for each study.}
}
{\FL
        &         &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
\NN
\cmidrule{4-5}  \cmidrule{6-7}
Dataset & Y-range\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
\ML
Cavalli & 7.21 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
        & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
        & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
%%
Costanzo & 288.03& $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
       & & $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
%%
Kalla & 88.25 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
      &  & $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
%%
Dai & 32.71 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
     &   & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
\LL
}

\ctable[cap={}, caption={Performance metrics for the random forest
  models built on the single molecule activity data, for each ChEMBL dataset. In each case the model
was built using the logarithm of the reported activities.},
label={tab:rfmodel-orig}]
{lrrr}
{

}
{\FL
Dataset & Y-range & RMSE & $R^2$
\ML
Costanzo & 4.67 & 0.80 & 0.18 \NN
Kall          & 4.44 & 1.00 & 0.32 \NN
Dai          &  3.60 & 0.76 & 0.34
\LL
}

>>>>

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small structure-activity
  relationship datasets). We extracted structure-activity data for several ChEMBL assays and
  developed random forest models to predict SALI values, from pairwise combinations of molecular
  descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
  significant cliffs tended to be better than on the lesser ones. While the models do not exhibit
  very high levels of accuracy, our results indicate that they are able to prioritize molecules in
  terms of their ability to activity cliffs, thus serving as a tool to prospectively identify
  activity cliffs..
\end{abstract}

@@

In either case, one can take the numerical values and visualize them in a variety of ways ranging
from heatmaps of SALI matrices to network representations \cite{Wawer:2008aa,Guha:2008aa}. These
visualizations then allow the user to explore the landscape, quickly identifying a range of activity
cliffs, which an then be examined in detail. Apart from identifying individual activity cliffs, a
variety of other SAR constructs, such as ``activity ridges'' \cite{Vogt:2011bs} and multi-target
landscapes\cite{Dimova:2011fk} can also be identified and characterized.

@@

\section{Results}
\label{sec:applications}
We first consider the application of the SALI prediction methodology to the Cavalli dataset. We
developed three random forest models, using the three aggregation descriptor aggregation functions
described in Section \ref{sec:methodology}. Fig.~\ref{fig:rfmodel-cavalli} presents the predicted
versus observed SALI values from the three models and Table \ref{tab:rfmodel-results} summarizes the
performance metrics for the three models. Overall the different aggregation functions do not differ
dramatically in terms of the final model performance as root mean square error (RMSE) on the test
set ranges from 1.01 to 1.07 log units, which is lower compared to the standard deviation of the
dependent variable. However, $f_{\textrm{mean}}$ appears to lead to the best performance (RMSE =
1.01 and $R^2$ = 0.58). In contrast, the RMSE and $R^2$ values for Y-scrambled\cite{Rucker:2007aa}
models using the different aggregation functions ranged from 1.58 to 1.61 log units and 0.003 to
0.01, respectively.

@@

We observed that the performance of models based on geometric mean aggregation function did not
differ substantially from those built with the arithmetic mean. This was also true of the other
datasets and so the following discussion omits the results obtained when using
$f_{\textrm{geom}}$. In fact, a paired $t$-test between the predicted SALI values (for the training
set) using models built with the different aggregation functions (i.e., $f_{\textrm{mean}}$ vs
$f_{\textrm{diff}}$, etc.)  indicated that the differences between the sets of predictions was
statistically insignificant (all $p$-values $> 0.1$). Fig.~\ref{fig:rfmodel-chembl} summarizes the
models built on the ChEMBL datasets using $f_{\textrm{diff}}$ and $f_{\textrm{mean}}$ as the
aggregation functions. Table \ref{tab:rfmodel-results} reports the model statistics Of the three,
the Costanzo stands out as being rather poor given the training and test set RMSE values being quite
close to (or greater than) the standard deviation of the dependent variable.  While the differences
in model performance do not vary significantly with the aggregation function used, we see that for
all the ChEMBL datasets, $f_{\textrm{diff}}$ leads to a slightly better model in terms of RMSE and
$R^2$ than when using $f_{\textrm{mean}}$. As with the Cavalli dataset, the SALI models exhibit more
variance for small SALI values (i.e., less significant cliffs) but appear to perform better at the
higher end of the SALI values. However in all three cases, systematic prediction errors are
apparent. Based on these observations, we arbitrarily chose $f_{\textrm{diff}}$ as the aggregation
function for subsequent analyses. Table \ref{tab:rfmodel-orig} and Figure \ref{fig:rfmodel-orig}
summarizes the performance of the models built using the reported activity data (original, single
molecule activity) for the Costanzo, Kalla and Dai datasets. Each of these models employed the
entire dataset (rather than splitting into a training and testing set) and exhibit relatively poor
performance as indicated by the RMSE (ranging from 0.76 to 1.00 log units) and $R^2$ (0.18 to
0.34). The model for the Kalla dataset is especially poor, given the RMSE is greater than the
standard deviation of the dependent variable. The poor performance of these models can be partially
ascribed to the small sizes of the dataset. It should be note that this methodology is not aimed at
modeling the original, usually small, SAR dataset. It is well known (and evident from the current
analyses) that small datasets do not necessarily lead to very robust or reliable machine learning
models. Instead, the methodology focuses on \emph{pairwise} datasets -- which are usually large
enough for a statistically significant model. Based on the poor performance observed for the
Costanzo dataset, we excluded it from our subsequent discussions.

@@

\ctable[cap={}, caption={Performance metrics for the random forest
  models built for the datasets used in this study. SD is the standard
deviation of the dependent variable.},
label={tab:rfmodel-results}]
{lrrcrrrr}
{
\tnote{Y-range is calculated using the entire dataset for each study.}
}
{\FL
        &    &     &                      & \multicolumn{2}{c}{Training} & \multicolumn{2}{c}{Test}
\NN
\cmidrule{5-6}  \cmidrule{7-8}
Dataset & Y-range\tmark & SD\tmark & Aggregation Function & RMSE & $R^2$ & RMSE & $R^2$   
\ML
Cavalli & 7.21 & 1.57 & $f_{\textrm{diff}}$ & 1.16& 0.54 & 1.07& 0.56\NN
        & & & $f_{\textrm{mean}}$      & 1.03& 0.64 & 1.01& 0.58\NN
        & & & $f_{\textrm{geom}}$      & 1.07& 0.61 & 1.01& 0.59\NN \NN
%%
Costanzo & 288.02& 14.84 & $f_{\textrm{diff}}$ & 11.99& 0.24 & 17.97& 0.10 \NN
       & & &  $f_{\textrm{mean}}$         & 12.88& 0.12 & 16.96& 0.20\NN \NN
%%
Kalla & 88.25 & 12.16 & $f_{\textrm{diff}}$ & 8.12& 0.59 & 6.49& 0.57\NN
      &  & &  $f_{\textrm{mean}}$      & 8.79& 0.52 & 8.00& 0.47\NN \NN
%%
Dai & 32.71 & 3.48 & $f_{\textrm{diff}}$ & 2.49& 0.48 & 3.21& 0.22\NN
     &   & & $f_{\textrm{mean}}$    & 2.64& 0.41 & 2.86& 0.37
\LL
}

\ctable[cap={}, caption={Performance metrics for the random forest
  models built on the single molecule activity data, for each ChEMBL dataset. In each case the model
was built using the $\log_{10}$ of the reported activities.},
label={tab:rfmodel-orig}]
{lrrrr}
{

}
{\FL
Dataset & Y-range & SD & RMSE & $R^2$
\ML
Costanzo & 4.67 & 1.22 & 0.80 & 0.18 \NN
Kalla          & 4.44 & 0.78 & 1.00 & 0.32 \NN
Dai          &  3.60 & 0.94 & 0.76 & 0.34
\LL
}
